---
title: "SFCC_Internship_Analyses"
author: "Hamish Williams"
date: "2024-07-04"
output: word_document
editor_options: 
  chunk_output_type: console
---

# Working Notes Document - Week 2 SFCC Internship

In this document I will go through chronologically the tasks completed and discuss the results of the work completed to date. This markdown specifically is to serve as a reproducible and good example of how a markdown document can be made to inform others on the work completed with accessible coding notations, and in the instance of this internship and the goals set out at the start of the internship, to generate a reproducible document which could be used to further assess data in the future.

There will be several parts to this markdown. I will include spaces for the necessary packages and custom functions used to generate the results, as well as directing the user to the appropriate spaces where the code for the shiny apps produced are stored (note: I will include the coding for the apps in here, but will likely have to comment these coding sections to render them nonfunctional, otherwise the generation of this file will not be possible). I will also give some notes on my thoughts and ideas from the interpretations of the results, and how these could be utilized by the SFCC and FMS in the future to better collate and analyse their costings/shortfalls for the fisheries boards/trusts, whilst considering the issues and costs of altering the data gathering processes.

## GitHub

All of the work completed here is accesible from the GitHub page which I have created to work on the data given for the internship: (<https://github.com/HamishGWilliams/SFCC_Internship.git>). For the sake of transparency and accessibility to the GitHub, I have made this public currently so that users may access the github and view the data, results and apps.

## Shiny Apps

Speaking of apps, there are currently 2 live, which can be accessed from here: The first is the initial shiny app made which allows an interactive session to play around with a boxplot figure representing the funding shortfalls variable from the raw data (<https://hamish-williams.shinyapps.io/SFCC_Internship_Interactable_App_1/>), the second app is more refined and has had serveral to a dozen processes completed to generate the necessary results to form the plots; This second shiny app provides and interactble session where the user and select between the differenct actions, scales and set a costings threshold on a sliding scale to show the boxplots of the costings of the various actions across all districts (<https://hamish-williams.shinyapps.io/SFCC_Internship_Interactable_App_2/>).

More information on how these were created and the modifications to the data completed to generate the results for these apps will be shown in a later part of this document

# Setup: loading packages and writing custom functions

The following are the neccessary packages required to complete the analyses proceeding, if you do not have these packages currently downloaded, then use the commented example to download the package(s) required on your device:

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_packages}
# Example of package installation
  # install.packages("ggplot2")
# install.packages("writexl")
# install.packages("openxlsx")
# install.packages("cowplot")

# Load the packages
library(dplyr)
library(ggplot2)
library(shiny)
library(here)
library(rsconnect)
library(writexl)
library(openxlsx)
library(cowplot)
library(gridExtra)
```

```{r custom_functions}
# Function to calculate means for a given scale
calculate_means <- function(data, scale) {
  data %>%
    group_by(Proposed.Management.Action) %>%
    summarize(average_cost = mean(Filtered_Costings_Data, na.rm = TRUE)) %>%
    mutate(scale = scale)
}

calculate_Funding.Shortfall_means_district <- function(data, District) {
  data %>%
    group_by(Proposed.Management.Action) %>%
    summarize(average_cost = mean(Funding.Shortfall, na.rm = TRUE)) %>%
    mutate(District = District)
}

```

# Excel work

A key issue exists with the raw excel data, which is that where and what the funding shortfall is based on is ambiguous. At times it seems to reflect the capital costings, the in-kind contributions, the annual costs, or even all 3 combined, whereas at other times it exists externally to the costings, and is what I would call a "True" shortfall; Which is to say that the shortfall given is not tied directly to the costings variables.

To compensate for this, an idea was proposed and reviewed by Sean, which is to devise a method to determine the likelihood that the funding shortfall is representative of the costings or not, and based on this combine all of the costings and shortfalls together to make a representative value for the total cost of the action.

The overall objective was to create a new data variable which instead of representing the funding shortfall, would represent a best estimate costs of the action of that row of data. Using this true cost, you could then subtract the funding shortfall from costings of this

There was a substantial amount of work completed not on this document, but on the excel spreadsheet (copy) of the data. I will take you through the steps completed to the excel data

## Key Considerations

From an email discussion with Sean, the following points were considered, quote:

1.  If its 0 I think they will not need funding -- there is another "Funding_Status" column, where this is listed as "unfunded" and the shortfall is 0 then I would interpret that as the action has not been costed for whatever reason. If the funding status is "funded" or "partially funded" and the shortfall is 0, I would interpret that as not needing extra cash.

2.  Yes, for those type of actions you have mentioned I think it would be appropriate to combine the capital/in-kind and annual costs to replace the 0 value in the "Funding Shortfall" column, as I know that is how some of the trusts calculated the shortfall. If a new field is created combining the shortfall and other costs then we may risk double-counting.

From this I have made the following assumptions:

-   If the funding shortfall is unfunded AND = 0, then don't include the shortfall in the calculation as it has not been costed for

-   If the funding shortfall is unfunded BUT \> 0, then include this in the calculations

-   If the shortfall is funded OR \> 0, then the shortfall could be included in the calculation

-   IF the funding shortfall is not a reflection of either the capital, in-kind or annual costs (i.e., shortfall = cost), then the shortfall should be included in the calculation

-   IF the funding shortfall is unfunded OR = 0, AND IS A reflection of any of the costings, then do not include the shortfall in the calculations

## Steps taken to generate new variable

The following steps were then followed to complete this set of criteria:

1.  Make a new sheet with a copy of all the data

2.  remove redundant variables not used in data, making sure to include the ID column in case there is need for one of the removed columns in the future

3.  New Column: "Funding_Status_TRUE" - identifies whether the action has funding or not.

-   ***=IF(OR(\$M2="Partially funded", \$M2="Funding secured"), "Funded", "Unfunded")***
-   where M column = "Funding Available
-   Key aim is to help discern if the funding shortfall is based on the costings variables or not, and was an advisement of Sean when asking for considerations for this objective

4.  New column: 'Shortfall_Capital' - identifies if shortfall reflects capital costings

-   ***=IF(AND(Q2\>0,Q2=I2),"Yes","No")***
-   Where Q = shortfall and I = capital costs

5.  New columns: "Shortfall_In-Kind" & "Shortfall_Annual" - same as before, but for the other costings

-   ***=IF(AND(Q2\>0,Q2=J2),"Yes","No")*** for the in-kind
-   ***=IF(AND(Q2\>0,Q2=L2),"Yes","No")*** for the annual

6.  New column: "Condition_total_costings" - uses the previously generated criteria to determine how if it calculates the costs including the shortfall or not:

-   ***=IF(AND(OR(R2="Funded",Q2\>0),S2="No",T2="No",U2="No"),SUM(I2,J2,L2,Q2),"")***
-   IF the criteria in the row shows that the action is funded OR has a shortfall \> 0, AND the shortfall does NOT reflect any of the costings, then it will sum the values of all the costings and the shortfall. IF NOT, then it will leave a blank cell
-   For brevities sake leaving a blank cell could be replaced with simply just calculating the costings together, but I did not do that before as I was getting confused with the excel coding. I instead made the combined costings in a new column

7.  New column: "Combined_Costings" - combines all the costing together

8.  New column: "Chossing the costing variable" - selects which calculation to take show

-   \*\*\*=IF(V2="",W2,V2)\*\*\* IF the Condition variable ="", then select the combined costings, otherwise use the conditioned variable.
-   Again, would not need this if I had tidied up the coding more, but its not much more space needed to complete the step, and separating these steps also means that Its easier to make changes to individual steps at a time.

9.  New column: "Filtered_Costings_Data" - removes any 0's and replaces them with "" (NAs) as to not have these values skew the costings averages towards lower estimates.

-   \*\*\*=IF(X2=0,"",X2)\*\*\* simply takes the value, IF = 0, then replace with""

I then saved this new sheet as "Fishery_Management_Plan_Actions_Working_Sheet.csv", which is then used in the analyses completed later.

This is not a perfect strategy, however I feel like it does a much better job to try and represent the actual costs of actions, but just the shortfalls. Using this new variable, we can then take the costings or the funding shortfalls and be able to calculate the aboslute and relative parts of the actions which are not accounted for with funding currently.

# R Work

The next step was to try and create some summary data of the costings, and to illustrate the distributions of the costings in a box plot

```{r import_data}
# Import Data
data <- read.csv("../data/Fishery_Management_Plan_Actions_Working_Sheet.csv", header = T)
wb <- createWorkbook()
addWorksheet(wb, "Sheet1")
writeData(wb, sheet = "Sheet1", x = data)
saveWorkbook(wb, "Fishery_Management_Plan_Actions_Working_Sheet_Data.xlsx", overwrite = TRUE)

# Change data type of relevant variables
str(data)
data$Proposed.Management.Action <- as.factor(data$Proposed.Management.Action)
data$Scale <- as.factor(data$Scale)
data$Proposed.Management.Action <- as.factor(data$Proposed.Management.Action)
data$Maintainance.Required <- as.factor(data$Maintainance.Required)
data$Funding.Available <- as.factor(data$Funding.Available)
data$Funding_Status_TRUE <- as.factor(data$Funding_Status_TRUE)
data$Maintainance.Required <- as.factor(data$Maintainance.Required)
data$Costing.Certainty <- as.factor(data$Costing.Certainty)
```

```{r Alter_data}
# Subset to very confident funding shortfall data
data_subset <- data[data$Costing.Certainty == "Very Confident",]
# Save this subsetted dataframe to a new file for use.
write.csv(data_subset, file = "../results/SFCC_VC_Subset.csv")

# subset data to relevant scales for investigation: 1-3
data_scale_1 <- data_subset[data_subset$Scale == "1",]
str(data_scale_1)
data_scale_2 <- data_subset[data_subset$Scale == "2",]
str(data_scale_2)
data_scale_3 <- data_subset[data_subset$Scale == "3",]
str(data_scale_3)
data_scale_4 <- data_subset[data_subset$Scale == "4",]
str(data_scale_4)
data_scale_5 <- data_subset[data_subset$Scale == "5",]
str(data_scale_5)

# for each scale, calculate the means of funding shortfall per district:

# Define the expanded function with checks for column existence
calculate_funding_shortfall_by_district_and_action <- function(data) {
  # Calculate the average values for Funding.Shortfall and Filtered_Costings_Data for each District and Proposed.Management_Action
  averages <- data %>%
    group_by(Proposed.Management.Action, District) %>%
    summarize(
      Average_Funding_Shortfall = mean(Funding.Shortfall, na.rm = TRUE),
      Average_Filtered_Costings_Data = mean(Filtered_Costings_Data, na.rm = TRUE),
      Count = n()
    )
  
  return(averages)
}
data_scale_1$Proposed.Management.Action
# Calculate the averages

# Scale 1
district_averages_scale_1 <- calculate_funding_shortfall_by_district_and_action(data_scale_1)
district_averages_scale_1$District <- as.factor(district_averages_scale_1$District)

par(mfrow = c(2,1))

plot(district_averages_scale_1$District, district_averages_scale_1$Average_Funding_Shortfall, 
     main = "Districts ~ Funding Shortfall (Scale 1)", 
     xlab = "", 
     ylab = "Average Funding Shortfall (£)",
     cex.axis = 0.7,
     las = 3,
     cex.main = 0.8)

plot(district_averages_scale_1$Proposed.Management.Action, district_averages_scale_1$Average_Funding_Shortfall, 
     main = "Actions ~ Funding Shortfall (Scale 1)", 
     xlab = "Proposed Management Actions", 
     ylab = "Average Funding Shortfall (£)",
     cex.axis = 0.7,
     cex.main = 0.8)

par(mfrow = c(1,1))

ggplot(district_averages_scale_1, aes(x = Proposed.Management.Action, y = Count)) +
  geom_bar(stat = "identity") +
  labs(title = "Count of Actions (Very Confident @ Scale = 1)",
       x = "Proposed Management Action",
       y = "Count") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 0)) +
  scale_x_discrete(drop = FALSE) +
  scale_fill_discrete(name = "Action")

ggplot(district_averages_scale_1, aes(x = District, y = Count)) +
  geom_bar(stat = "identity") +
  labs(title = "Count of Actions for per district (Very Confident @ Scale = 1)",
       x = "District",
       y = "Count") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_discrete(drop = FALSE) +
  scale_fill_discrete(name = "Action")


# Looks like the Forth has a very high average funding shortfall compared to others
  # built from 2 separate actions 

# Scale 2
district_averages_scale_2 <- calculate_funding_shortfall_by_district_and_action(data_scale_2)

district_averages_scale_2$District <- as.factor(district_averages_scale_2$District)

plot(district_averages_scale_2$Proposed.Management.Action, district_averages_scale_2$Average_Funding_Shortfall)

plot(district_averages_scale_2$District, district_averages_scale_2$Average_Funding_Shortfall)
# Very few action points @ scale 2

# Scale 3
district_averages_scale_3 <- calculate_funding_shortfall_by_district_and_action(data_scale_3)

district_averages_scale_3$District <- as.factor(district_averages_scale_3$District)

par(mfrow = c(2,1))
plot(district_averages_scale_3$Proposed.Management.Action, district_averages_scale_3$Average_Funding_Shortfall, main = "Actions ~ Funding Shortfall (Scale 3)", xlab = "Proposed Management Actions", ylab = "Average Funding Shortfall (£)")

plot(district_averages_scale_3$District, district_averages_scale_3$Average_Funding_Shortfall, main = "Districts ~ Funding Shortfall (Scale 3)", xlab = "Districts", ylab = "Average Funding Shortfall (£)")

par(mfrow = c(1,1))

# SCale 4
district_averages_scale_4 <- calculate_funding_shortfall_by_district_and_action(data_scale_4)
district_averages_scale_5 <- calculate_funding_shortfall_by_district_and_action(data_scale_5)

#  Write all of the new datafiles to objects:
getwd()

write.csv(district_averages_scale_1, "./data/district_averages_scale_1.csv")
write.csv(district_averages_scale_2, "./data/district_averages_scale_2.csv")
write.csv(district_averages_scale_3, "./data/district_averages_scale_3.csv")
write.csv(district_averages_scale_4, "./data/district_averages_scale_4.csv")
write.csv(district_averages_scale_5, "./data/district_averages_scale_5.csv")

```

```{r boxplots_1}
# Boxplot of filtered costings data ~ actions
plot(data_subset$Proposed.Management.Action, data_subset$Filtered_Costings_Data)

# ggplot
ggplot(data_subset, aes(x = Proposed.Management.Action, y = Filtered_Costings_Data, fill = Proposed.Management.Action)) +
         geom_boxplot() + 
         facet_wrap(~Scale, scales = "free") + 
  labs(title = "Average Cost by Proposed Management Action for all Scales",
       x = "Proposed Management Action",
       y = "Average Cost") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_discrete(name = "Action")
```

```{r generating_new_average_dataset}
# Investigating the means of each scale for their costs per action group:
by(data_scale_1$Filtered_Costings_Data, data_scale_1$Proposed.Management.Action, summary)
by(data_scale_2$Filtered_Costings_Data, data_scale_2$Proposed.Management.Action, summary)
by(data_scale_3$Filtered_Costings_Data, data_scale_3$Proposed.Management.Action, summary)
by(data_scale_4$Filtered_Costings_Data, data_scale_4$Proposed.Management.Action, summary)
by(data_scale_5$Filtered_Costings_Data, data_scale_5$Proposed.Management.Action, summary)

  # Initial observations tell us that not all scales have points on all management actions.
  # We can also see instances where there are very few datapoints for some actions at given scales
  # another observation is that the mean and median values of actions or scales are not equal, which shows bias towards either the lower or upper values no the distrbution curves (assuming normall ditributed). This is likely not a concern, but something to consider, that some of the actions may have some biasing datapoints which pull the data in a particular direction
    # Another pointto add to the last, the distributions are limited to a mimimum value of 0 (which in this instance has not been included as detailed in the excel section).

# Function to calculate means for a given scale
calculate_means <- function(data, scale) {
  data %>%
    group_by(Proposed.Management.Action) %>%
    summarize(average_cost = mean(Filtered_Costings_Data, na.rm = TRUE)) %>%
    mutate(scale = scale)
}

# Calculate means for each scale
results_scale_1 <- calculate_means(data_scale_1, "Scale 1")
results_scale_2 <- calculate_means(data_scale_2, "Scale 2")
results_scale_3 <- calculate_means(data_scale_3, "Scale 3")
results_scale_4 <- calculate_means(data_scale_4, "Scale 4")
results_scale_5 <- calculate_means(data_scale_5, "Scale 5")

# Combine results into a single data frame
results <- bind_rows(results_scale_1, results_scale_2, results_scale_3, results_scale_4, results_scale_5)

# Ensure all actions (A:T) are represented for each scale
all_actions <- expand.grid(
  scale = unique(results$scale),
  Proposed.Management.Action = factor(LETTERS[1:20])
)

# Merge to ensure all actions are present, filling with NA where no data is available
results_full <- merge(all_actions, results, by = c("scale", "Proposed.Management.Action"), all.x = TRUE)
write.csv(results_full, file = "../results/average_spending_results.csv")

wb <- createWorkbook()
addWorksheet(wb, "Sheet1")
writeData(wb, sheet = "Sheet1", x = results_full)
saveWorkbook(wb, "./results/average_spending_results.xlsx", overwrite = TRUE)
```

```{r plot_data}
# Barplot of average results
barplot(results_scale_1$average_cost~results_scale_1$Proposed.Management.Action,
        xlab = "Action", ylab = "Cost (£)", main = "Average cost of action (scale 1)")

ggplot(results_full, aes(x = Proposed.Management.Action, y = average_cost, fill = Proposed.Management.Action)) +
  geom_bar(stat = "identity") +
  labs(title = "Average Cost by Proposed Management Action for all Scales",
       x = "Proposed Management Action",
       y = "Average Cost") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 0)) +
  scale_fill_discrete(name = "Action") + 
  facet_wrap(~scale, scales = "free_y", nrow = 5)

# Boxplot of filtered costings data ~ actions
ggplot(data_subset, aes(x = Proposed.Management.Action, y = Filtered_Costings_Data, fill = Proposed.Management.Action)) +
         geom_boxplot() + 
  labs(title = "Costs by Proposed Management Action for all Scales",
       x = "Proposed Management Action",
       y = "Cost (£)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 0)) +
  scale_fill_discrete(name = "Action") + 
  facet_wrap(~Scale, scales = "free_y", nrow = 5) 
```

These plots are useful and provide a good basis to complete future work for refinement and presentation, and are submitted as such in this document.

Looking at the box plot, we can see, like mentioned in the comments of one of the last coding chunks, that there are some actions of different scales which are skewing the data quite a lot (represented here by outlier points on the boxplots). These limit our ability to see the shapes and distributions of the boxplots, however we should not throw these values to the way-side just yet, as they likely do represent real costs to specific actions.

What might be or more value to these plots is to further investigate the distribution of the costings across the districts and to see if these high values are all tied to a specific district, which might have certain characteristics and circumstances where these actions cost a great deal more for these specific actions on these districts.

Another observation is that there are a number of actions with little to no data points. Which is to say, that it is difficult to determine what their "likely average" costs are due to insufficient data present at the time. Given the fact that there is quite a range of costs for any given action (at each scale and also at each confidence - which is not shown at this time), which gives me cause for concern about trying to apply average values from these low sample data variables to provide "better" estimates for actions with lower estimate confidences.

That isn't to say that these CANT be utilized, but if they are, they should be done with the consideration that there is possibilities that they may not be any more effective at estimate the total costs of a given action than professional opinion from contractors or inside personalle with experience in the managements, executions and therefore the costings of management actions.

Notes from meeting (05/07/24):

\* look at how different districts "use" the funding shortfall variable

\* Add in selections or costing confidences for 2nd shiny app, and maybe even district

\* Focus more next week on the funding shortfall variable, don't worry too much about what and how it represents the funding shortfall

\* Look at back-calculating the 0 values in the funding shortfalls

```         
- considerations for instances where there isn't data for the shortfall \@ the scale, do I use different scale values?

- Create a range (+ £10,000 from average estimate to top value)

- Make a note of which actions you feel more, less, not confident about the average values

- make the confidence based on the count of the actions used to calculate the average
```

\* Tuesday field-day

```         
- Pamala & Sam

- sT PETERS HERITAGE CENTRE - Peterculter

    - Look for buses to Peterculter

- 9:30am meet-up at the heritage centre

- Full day, 4 sites, bring a lunch and water
```

# Part 2 - Focusing on Funding Shortfall

So initially I was working a lot on trying to create a new variable as I was coming to grips with understanding what exactly the "Funding shortfall" and "costings" variables were actually representative of.

I made a shiny app to run this (which still seems to not be working at this time for some reason - and I'm probably unlikely to come back to this app since I have just developed a better one.). However, when I had a discussion with sean at our latest meeting, we both seemed to have issues understanding what the funding shortfall actually represents, and it seemed that my understanding that it could be reflective of 2 different scenarios (but not specified which it could be) may have been a misunderstanding, and that the issue I was raising was not of precident at the moment. 

We decided that I would focus my efforts back to just the funding shortfall (trying not too hard to worry about it still!), and to come up with a table or set of results which could represent the average values for actions at very confident status at the different scales - initially focusing on scale 1, but still allowing some room to consider the other scales. This would also include work to understand how the districts played a part in these funding shortfalls, and if any specific district had tended to have relatively high funding shortfalls compared to others.

The main goals thus have become:
  - To explore the funding shortfall variable
  - To create a set of results which contain the average funding shortfalls for the different actions at the different scales
  - To understand how representative these average values are of the whole distribution
  To generate a new Shiny app which could be used to interactively display and visually show the data.
  
Lets start by re-loading the data we used, just to help with flow and understanding the process that was undertaken:

```{r Part_2_Import_Data}
# Import Data
data <- read.csv("../data/Fishery_Management_Plan_Actions_Working_Sheet.csv", header = T)
wb <- createWorkbook()
addWorksheet(wb, "Sheet1")
writeData(wb, sheet = "Sheet1", x = data)
saveWorkbook(wb, "Fishery_Management_Plan_Actions_Working_Sheet_Data.xlsx", overwrite = TRUE)

# Change data type of relevant variables
str(data)
data$Proposed.Management.Action <- as.factor(data$Proposed.Management.Action)
data$Scale <- as.factor(data$Scale)
data$Proposed.Management.Action <- as.factor(data$Proposed.Management.Action)
data$Maintainance.Required <- as.factor(data$Maintainance.Required)
data$Funding.Available <- as.factor(data$Funding.Available)
data$Funding_Status_TRUE <- as.factor(data$Funding_Status_TRUE)
data$Maintainance.Required <- as.factor(data$Maintainance.Required)
data$Costing.Certainty <- as.factor(data$Costing.Certainty)
```

Now that  have imported the data again, we are going to subset the data to only include values which are "Very Confident":

```{r Very_Confident_Subset}
# Subset to very confident funding shortfall data
data_subset <- data[data$Costing.Certainty == "Very Confident",]
# Save this subsetted dataframe to a new file for use.
write.csv(data_subset, file = "../results/SFCC_VC_Subset.csv")
```

We save the data here just because it might be useful to have a separate data file which is only the very confident data.

Now I'll look to create separate datafiles which contain only the data for the different levels of scale. If I want to go back get the other confidence intervals then I can, but since we are only concerning ourselves with the very confident values for now, then I'm going to subset for that. I'm separating them based on scale to make it easier to work with the data later in the shiny app, as its relatively easy enough to merge the rows if needed at a later point:

```{r Subsetting_Scales}
# subset data to relevant scales for investigation: 
data_scale_1 <- data_subset[data_subset$Scale == "1",]
str(data_scale_1)
data_scale_2 <- data_subset[data_subset$Scale == "2",]
str(data_scale_2)
data_scale_3 <- data_subset[data_subset$Scale == "3",]
str(data_scale_3)
data_scale_4 <- data_subset[data_subset$Scale == "4",]
str(data_scale_4)
data_scale_5 <- data_subset[data_subset$Scale == "5",]
str(data_scale_5)
```

We can then perform some functions in order to calculate the average values for funding shortfall based on the district and actions

```{r calculating_means}
## Define the expanded function with checks for column existence
calculate_funding_shortfall_by_district_and_action <- function(data) {
  # Calculate the average values for Funding.Shortfall and Filtered_Costings_Data for each District and Proposed.Management_Action
  averages <- data %>%
    group_by(Proposed.Management.Action, District) %>%
    summarize(
      Average_Funding_Shortfall = mean(Funding.Shortfall, na.rm = TRUE), # Average Funding Shortall
      Average_Filtered_Costings_Data = mean(Filtered_Costings_Data, na.rm = TRUE), # Average the costings data I made
      Count = n() # calculate the number of values used to generate the average
    )
  return(averages)
}

## Calculate the averages
# Scale 1
district_averages_scale_1 <- calculate_funding_shortfall_by_district_and_action(data_scale_1)
district_averages_scale_1$District <- as.factor(district_averages_scale_1$District) # Need to change this to factor since it was causing issues with creating plots
# Scale 2
district_averages_scale_2 <- calculate_funding_shortfall_by_district_and_action(data_scale_2)
district_averages_scale_2$District <- as.factor(district_averages_scale_2$District)
# Scale 3
district_averages_scale_3 <- calculate_funding_shortfall_by_district_and_action(data_scale_3)
district_averages_scale_3$District <- as.factor(district_averages_scale_3$District)
# Scale 4
district_averages_scale_4 <- calculate_funding_shortfall_by_district_and_action(data_scale_4)
# Scale 5
district_averages_scale_5 <- calculate_funding_shortfall_by_district_and_action(data_scale_5)

## Write the files and save them for use in app:
write.csv(district_averages_scale_1, "./data/district_averages_scale_1.csv")
write.csv(district_averages_scale_2, "./data/district_averages_scale_2.csv")
write.csv(district_averages_scale_3, "./data/district_averages_scale_3.csv")
write.csv(district_averages_scale_4, "./data/district_averages_scale_4.csv")
write.csv(district_averages_scale_5, "./data/district_averages_scale_5.csv")
```

```{r Plotting_Averages}
# Scale 1
par(mfrow = c(2,1))

plot(district_averages_scale_1$District, district_averages_scale_1$Average_Funding_Shortfall, 
     main = "Districts ~ Funding Shortfall (Scale 1)", 
     xlab = "", 
     ylab = "Average Funding Shortfall (£)",
     cex.axis = 0.7,
     las = 3,
     cex.main = 0.8)

plot(district_averages_scale_1$Proposed.Management.Action, district_averages_scale_1$Average_Funding_Shortfall, 
     main = "Actions ~ Funding Shortfall (Scale 1)", 
     xlab = "Proposed Management Actions", 
     ylab = "Average Funding Shortfall (£)",
     cex.axis = 0.7,
     cex.main = 0.8)

par(mfrow = c(1,1))

## These plots ar quite nice, I like them a lot and they provide a lot of information that is useful. We can clearly see that the Forth district is responsible for the 2 largest shortfall values in the bottom plot, and that all the others seem to range from 0 - £50,000 at the highest points. What is causing this large shortfall in the Forth is unknown at this time, but should be investigated further. It seems isolated to actions D and B, so perhaps the expanded definitions of these actions could provide insight into how they contribute to the significantly higher shortfall values.
```

Although the plots above are useful and will be used, an issue remains that we don't know how representative these boxes are for each action @ scale & @ Very confident status. That is why I included a calculation for counting the number of values (rows) used to calculate the averages for each action at each district, so that we can better understand how they represent the "real" average values of the funding shortfalls.

```{r barplot_of_counts}
## Barplots for both the actions and the districts
# Build plots
district_averages_scale_1_action_barplot <- ggplot(district_averages_scale_1, aes(x = Proposed.Management.Action, y = Count)) +
  geom_bar(stat = "identity") +
  labs(title = "Count of Actions (Very Confident @ Scale = 1)",
       x = "Proposed Management Action",
       y = "Count") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 0)) +
  scale_x_discrete(drop = FALSE) +
  scale_fill_discrete(name = "Action")

district_averages_scale_1_district_barplot <- ggplot(district_averages_scale_1, aes(x = District, y = Count)) +
  geom_bar(stat = "identity") +
  labs(title = "Count of Actions for per district (Very Confident @ Scale = 1)",
       x = "District",
       y = "Count") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_discrete(drop = FALSE) +
  scale_fill_discrete(name = "Action")

# Arrange and pint plots
grid.arrange(district_averages_scale_1_action_barplot,
             district_averages_scale_1_district_barplot,
             nrow = 2)
```

What might be more uesful however, is to visualise how the barplots and boxplots look side by side for both action and districts:

```{r}
## Plot boxplot and barplots together
# Actions
action_boxplot_scale_1 <- ggplot(district_averages_scale_1, 
       aes(x = Proposed.Management.Action,
           y = Average_Funding_Shortfall)) + 
  geom_boxplot() + 
    labs(title = "Average FUnding Shortfall per Action",
       x = "Proposed Management Action",
       y = "Funding Shortfall (£)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 0)) +
  scale_x_discrete(drop = FALSE)

district_averages_scale_1_action_barplot <- ggplot(district_averages_scale_1, aes(x = Proposed.Management.Action, y = Count)) +
  geom_bar(stat = "identity") +
  labs(title = "Count of Actions (Very Confident @ Scale = 1)",
       x = "Proposed Management Action",
       y = "Count") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 0)) +
  scale_x_discrete(drop = FALSE) +
  scale_fill_discrete(name = "Action")

grid.arrange(action_boxplot_scale_1,
             district_averages_scale_1_action_barplot,
             nrow = 2)

## We can clearly see some actions which do not have many (or no) data for them at all for the scale, confidence and action. It would be pretty safe to say that for action B, that we can be fairly confident that the average value of the shortfall is representative aside from the one outlier which exists from the Forth district. Some of the other actions with potentiallt fairly representative averages are F, L, G & Q. 

## This really helps when trying to consider how representative the averages are to the actions, and highlights that in many cases, we don't have a lot of data to make a fair assessment that the average will be representative for he purposes it is used for.

## With this alone, we can start to summise that its likely that some action points cannot have a fair back-calculation of what the shortfalls will be, since we lack the data to assuringly say so.

## Another point which I want to observe about the purpose of the funding shortfall in this internship, is that not all individual actions in the same category will have the same shortfall, as some districts might already have access to better funding to others. Thus, I would reiterate that it would be better to get an average costing of a given action point in its totality (including capital and in-kind costs), with a secondary costings to account for the the potential maintenance costs which will occur annually. With this, we could better provide districts with estimates of what an action would cost. 

## HOweverm if the interest is to understand which actions and districts possess the greatest shortfalls, or are lacking in funding consistently, and to what extent, then the purposes of these analyses and better justified.

### NOTE: The comments given above are for primarily my opinions and thoughts as I come to understand what the potential use cases and purposes of my analyses would be applicable to, and are not reflections of me trying to undermine the purpose of the work. 

# Districts
district_boxplot_scale_1 <- ggplot(district_averages_scale_1, 
       aes(x = District,
           y = Average_Funding_Shortfall)) + 
  geom_boxplot() + 
    labs(title = "Average Funding Shortfall per District",
       x = "District",
       y = "Funding Shortfall (£)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_discrete(drop = FALSE)

district_averages_scale_1_district_barplot <- ggplot(district_averages_scale_1, aes(x = District, y = Count)) +
  geom_bar(stat = "identity") +
  labs(title = "Count of Actions for per district",
       x = "District",
       y = "Count") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_discrete(drop = FALSE) +
  scale_fill_discrete(name = "Action")

grid.arrange(district_boxplot_scale_1,
             district_averages_scale_1_district_barplot,
             nrow = 2)

## From these figures we can see that there are a lot of actions given to the Girvan, Irvine & Garnock, and the Stinchar districts, of which actions I am not sure at this moment, I can create an independent plot for this if it is of value.

ggplot(district_averages_scale_1, aes(x = District, y = Count)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_identity() + 
  labs(title = "Count of Actions for per district",
       x = "District",
       y = "Count") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~Proposed.Management.Action, scale = "free_x")

## Plotting this, we get a better idea of what maes up each district in terms of their actions @ scale 1 & @ very confident assessments of funding shortfalls. We can wee that action be has the most diversity in terms of districts, whith many other actions only being found in a few districts. Irvine & Garnock seem to have the most instances of action B @ present, and then Girvan and stinchar posessing the most action points for action Q.

  data_summary <- district_averages_scale_1 %>%
    group_by(Proposed.Management.Action, District) %>%
    summarise(Total_Count = sum(Count, na.rm = TRUE))

data_summary <- data_summary %>%
    arrange(desc(Proposed.Management.Action)) %>%
    mutate(prop = Total_Count / sum(Total_Count) * 100) %>%
    mutate(ypos = cumsum(prop) - 0.5 * prop)

ggplot(data_summary, aes(x = "", y = prop, col = Proposed.Management.Action)) +
    geom_bar(stat = "identity", width = 1) +
    coord_polar("y", start = 0) +
    scale_fill_identity() +
    theme_void() +
    labs(
      title = "Pie Chart by Proposed Management Action"
    ) +
    theme(legend.position = "bottom") +
    geom_text(aes(y = ypos, label = paste0(Proposed.Management.Action, ": ", round(prop, 1), "%")), color = "white")

```
  


1. Focus only on the funding shortfall variable, 
  - create a table which shows the average shortfalls of actions @ very confidence & @ each scale (1:3)
  
```{r import data}
getwd()
data <- read.csv("./data/Fishery_Management_Plan_Actions_Working_Sheet.csv", header = T)
# Change data type of relevant variables
str(data)
data$Proposed.Management.Action <- as.factor(data$Proposed.Management.Action)
data$Scale <- as.factor(data$Scale)
data$Proposed.Management.Action <- as.factor(data$Proposed.Management.Action)
data$Maintainance.Required <- as.factor(data$Maintainance.Required)
data$Funding.Available <- as.factor(data$Funding.Available)
data$Funding_Status_TRUE <- as.factor(data$Funding_Status_TRUE)
data$Maintainance.Required <- as.factor(data$Maintainance.Required)
data$Costing.Certainty <- as.factor(data$Costing.Certainty)


# Subset to very confident funding shortfall data
data_subset <- data[data$Costing.Certainty == "Very Confident",]
# Save this subsetted dataframe to a new file for use.
write.csv(data_subset, file = "./results/SFCC_VC_Subset.csv")

# Boxplot of funding shortfalls by scales
plot(data_subset$Proposed.Management.Action, data_subset$Funding.Shortfall)

# ggplot
ggplot(data_subset, aes(x = Proposed.Management.Action, y = Funding.Shortfall, fill = Proposed.Management.Action)) +
         geom_boxplot() + 
         facet_wrap(~Scale, scales = "free") + 
  labs(title = "Average Cost by Proposed Management Action for all Scales",
       x = "Proposed Management Action",
       y = "Average Funding Shortfall Cost") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_discrete(name = "Action")


# subset data to relevant scales for investigation: 1-3
data_scale_1 <- data_subset[data_subset$Scale == "1",]
str(data_scale_1)
data_scale_2 <- data_subset[data_subset$Scale == "2",]
str(data_scale_2)
data_scale_3 <- data_subset[data_subset$Scale == "3",]
str(data_scale_3)
data_scale_4 <- data_subset[data_subset$Scale == "4",]
str(data_scale_4)
data_scale_5 <- data_subset[data_subset$Scale == "5",]
str(data_scale_5)

# Create a boxplot which focuses only on scale 1:

ggplot(data = data_subset[data_subset$Scale == "1",], aes(x = Proposed.Management.Action, y = Funding.Shortfall, fill = Proposed.Management.Action)) +
         geom_boxplot() + 
         facet_wrap(~Scale, scales = "free") + 
  labs(title = "Average Cost by Proposed Management Action for all Scales",
       x = "Proposed Management Action",
       y = "Average Funding Shortfall Cost") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_discrete(name = "Action")


# New subset:

data_VC_scale_1 <- data_subset[data_subset$Scale == "1",]


hist(data_VC_scale_1$Funding.Shortfall, breaks = 20)
```

```{r}
ggplot(district_averages_scale_1, 
       aes(x = Proposed.Management.Action,
           y = Average_Funding_Shortfall,
           fill = District,
           col = District)) + 
  geom_boxplot() + 
    labs(title = "Average FUnding Shortfall per Action",
       x = "Proposed Management Action",
       y = "Funding Shortfall (£)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 0)) +
  scale_x_discrete(drop = FALSE)
```

